---
title: "Simulation Recreation"
author: "Hillary Heiling"
date: "September 8, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Recreation of one of the simulations in Section 2 of Raftery et al. (Estimating the integrating likelihood via posterior simulation using the harmonic mean identity)

Description of simulation: We are interested in the marginal likelihood for a single data point y

$$ y \sim N(\mu, \frac{1}{\psi}) $$

$$ \psi \sim Gamma(\alpha / 2, \alpha / 2) $$

Paper prior specification:
$$ \mu|\psi \sim N(\mu_0, n_0 \psi) $$

Correction to make things conjugate (I think), as the paper claims these priors are:

$$ \mu|\psi \sim N(\mu_0, n_0 \psi^{-1}) $$

Integrated likelihood analytical result:

$$ f(y) = St(y|\mu_0,\lambda = n_0/(n_0+1),\alpha) = \frac{\Gamma(\frac{1}{2}(\alpha+1))}{\Gamma(\frac{1}{2}\alpha)} \left ( \frac{\lambda}{\alpha \pi} \right)^{1/2} * [1 + \alpha^{-1} \lambda (y-\mu_0)^2]^{-(\alpha+1)/2} $$

Let the hyperparameters have the following values:

$ \mu_0 = 0, \alpha = 2, n_0 = 1 $

Target value:

```{r}
alpha = 2
mu = 0
y = 5
n0 = 1
lambda = n0 / (1+n0)

(f_y = gamma(0.5*(alpha+1)) / gamma(0.5*alpha) * (lambda / alpha / pi)^0.5 * 
  (1 + 1/alpha*lambda*(y-mu)^2)^((-1)*(alpha+1)/2))

# In harmonic mean functions, will shoot for following target value:
1/f_y

```

Note: Raftery et al. also used the following combinations: y = {5,3,0}, alpha = {2,6,10}, $\mu_0$ remained equal to 0.

## Gathering the MCMC samples

Gibbs approach: Sample $\psi$ from appropriate gamma dist and then sample $\mu$ based on given $\psi$ sample. ($\psi$ does not depend on $\mu$, so can independently sample $\psi$ from the specified gamma dist).

$$ \psi^{-1}|y \sim IG(\alpha/2, \alpha/2) \rightarrow \psi|y \sim Gamma(\alpha/2,\alpha/2) $$

Note: In general for multiple data points:

$$ \psi|y_1,...,y_n \sim Gamma(\alpha/2 + n/2 - 1/2, \alpha/2 + 1/2 \sum (y_i - \bar y)^2) $$
$$ \mu | \psi,y \sim N \left (\frac{y}{2}, \frac{\psi^{-1}}{2} \right) $$

Note: In general for multiple data points:

$$ \mu | \psi,y \sim N \left (\frac{n \bar y + \mu_0}{n + 1}, \frac{\psi^{-1}}{n + 1} \right) $$

```{r}
draws = function(mu0, alpha, M, y){
  
  psi = rgamma(M, shape = alpha / 2, scale = alpha / 2)
  mu = rnorm(M, mean = y/2, sd = sqrt(1/psi/2))
  
  return(cbind(mu,psi))
}
```


## Evaluating the Modified and Regular Harmonic Mean estimates using the MCMC samples

Once we have the MCMC samples, we can plug them into the (modified) harmonic mean functions. Returned value should be near target value given above.

```{r}
HM = function(MCMC, y){
  mu = MCMC[,1]
  psi = MCMC[,2] # precision of y; inverse of variance of y
  sd_y = sqrt(1/psi)
  dens = dnorm(y, mean = mu, sd = sd_y)
  inv_dens = 1 / dens
  
  # marg_lik = 1 / mean(inv_dens)
  # ll = log(marg_lik)
  
  return(mean(inv_dens))
}

mod_HM = function(MCMC, y, mu0, alpha, p){
  mu = MCMC[,1]
  psi = MCMC[,2]
  sd_y = sqrt(1/psi)
  
  # Calculation of denominator 
  dens = dnorm(y, mean = mu, sd = sd_y)
  prior = dnorm(mu, mean = mu0, sd = sqrt(1/psi)) * dgamma(psi, shape = alpha/2, scale = alpha/2)
  
  # Calculation of numerator
  # Posterior modes and covariance:
  mode = colMeans(MCMC)
  sigma = cov(MCMC)
  # Calculation of truncated normal:
  wt_norm = dmvnorm(MCMC, mean = mode, sigma = sigma)
  region = apply(MCMC, 1, function(x) t(x - mode) %*% solve(sigma) %*% (x - mode))
  cutoff = qchisq(p = p, df = 2) # df = number parameters, mu and psi
  trunc_norm = ifelse(region > cutoff, 0, wt_norm)
  print("proportion truncated:")
  print(sum(region > cutoff) / length(region))
  
  if(anyNA(trunc_norm)) print(sum(is.na(trunc_norm)))
  
  components = trunc_norm / (dens * prior)
  inv_lik = mean(components)
  # ll = log(1/inv_lik)
  
  return(inv_lik)
}
```

Results:

```{r}
library(mvtnorm)
mcmc_samps = draws(mu0 = 0, alpha = 2, 10^4, y = 5)

HM(mcmc_samps, 0)
mod_HM(mcmc_samps, y = 5, mu0 = 0, alpha = 2, p = 0.98)
```

The End